Supplemental Materials for ``Popularity Adjusted Block Models are Generalized Random Dot Product Graphs''

### 2022-02-06 referee-response.tex

\item
  p4. What is a hollow graph?\\
  \textcolor{blue}{
  We changed the terminology a bit here and throughout the paper and
  specify that the analysis assumes $G$ is an unweighted and undirected
  graph with no self-loops 
  (no edge directly from itself to itself), 
  so $A$ is binary and symmetric with zeros on the diagonal. The phrase
  ``hollow graph'' is no longer used.  }

@@

  p4. \(G\) as a graph is non-oriented or undirected and \(A\) is
  symmetric.\\
  \textcolor{blue}{
  We have made minor edits clarifying the structure of $G$ and $A$. In
  particular we assume throughout the paper that $G$ is undirected and $A$ is
  symmetric. 

@@

\item
  p4. Please define the set \([n]\).\\
  \textcolor{blue}{
  We added $[n] = \{1, 2, ..., n\}$ at the beginning of Section~2.1. 

@@

  \(z_1, ..., z_n\)? Please give the range of variation of \((i, j)\)
  when you define \(P_{ij}\).\\
  \textcolor{blue}{
  The range of $(i, j)$ is now included in the definition; in particular
  we wrote in Section 2.1 that ``$1 \leq i < j \leq n$.''
  In regard to the probability distribution on the labels $z_1, ...,
  z_n$, none of our theoretical assume anything about the distribution
  of the $\{z_i\}$; we do, however, occasionally assume that the $n_k$
  are sufficiently large for each $k = 1,2,\dots, K$. More specifically 
  the convergence rate of Theorem 5 depends on the $n_k$, i.e., for
  convergence we require that $n_k \to \infty$ as $n \rightarrow
  \infty$. In practice, we also require that each $n_k$ is large enough
  so that the latent vectors in the $k$th community span its
  subspace; this is a rather mild requirement, e.g., if the latent vectors in the $k$th community lies in a
  general position (are linearly independent), then we only require $n_k
  \geq K$.}

@@

  \(\mathcal{X}\) a subset on \((\mathbb{R}^d)^2\)? Please clearly state
  that \(d = p + q\) (not stated before Definition 3).\\
  \textcolor{blue}{
  We now denote $d = p + q$ in Definition 2 and define $\mathcal{X}$ as a subset of $\mathbb{R}^{d}$ such
  that for any $x \in \mathcal{X}$ and $y \in \mathcal{X}$, we have
  $x^{\top} I_{p,q} y \in [0,1]$. }

### 2022-02-06 sparsity-sim/supplemental-materials.tex

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames*,x11names*}{xcolor}
%
\documentclass[12pt]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\DeclareMathOperator*{\argmax}{arg\,max}
\DeclareMathOperator*{\argmin}{arg\,min}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Connecting the Popularity Adjusted Block Model to the Generalized Random Dot Product Graph for Community Detection and Parameter Estimation},
  colorlinks=true,
  linkcolor=Maroon,
  filecolor=Maroon,
  citecolor=Blue,
  urlcolor=blue,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
%\usepackage[margin=1in]{geometry}
\usepackage{graphicx,grffile}
\def\spacingset#1{\renewcommand{\baselinestretch}%
{#1}\small\normalsize} \spacingset{1}

\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\usepackage{float}
\usepackage{mathtools}
% \usepackage{natbib}
\usepackage[backend=bibtex,maxbibnames = 10,style=authoryear,giveninits=true,natbib]{biblatex}
\DeclareNameAlias{sortname}{last-first}
\addbibresource{misc.bib}
\renewbibmacro{in:}{%
  \ifentrytype{article}{}{%
  \printtext{\bibstring{in}\intitlepunct}}}
\usepackage{parskip}
\usepackage[linesnumbered,ruled,vlined]{algorithm2e}
% \setcitestyle{numbers,square,comma}
\usepackage{verbatim}
\usepackage{amsthm}
\usepackage{comment}
\addtolength{\oddsidemargin}{-.5in}%
\addtolength{\evensidemargin}{-.5in}%
\addtolength{\textwidth}{1in}%
\addtolength{\textheight}{1.3in}%
\addtolength{\topmargin}{-.8in}%

\title{Popularity Adjusted Block Models are Generalized Random Dot Product Graphs}

\author{John Koo\\Department of Statistics, Indiana University\\
Minh Tang\\Department of Statistics, North Carolina State University\\
Michael Trosset\\Department of Statistics, Indiana University}

% \date{\vspace{-2.5em}}
\date{}

\begin{document}

\maketitle

\begin{abstract}
We connect two random graph models, the Popularity Adjusted Block Model
(PABM) and the Generalized Random Dot Product Graph (GRDPG),
by demonstrating that a PABM is a GRDPG in which communities correspond to 
mutually orthogonal subspaces of latent vectors. 
This insight allows us to construct new algorithms for community detection
and parameter estimation for the PABM, 
as well as improve an existing algorithm 
that relies on Sparse Subspace Clustering. 
Using established asymptotic properties of 
Adjacency Spectral Embedding for the GRDPG, 
we derive asymptotic properties of these algorithms. 
In particular, we demonstrate that the absolute number of 
community detection errors tends to zero as 
the number of graph vertices tends to infinity. 
Simulation experiments illustrate these properties. 
\end{abstract}

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\newcommand{\diag}{\mathrm{diag}}
\newcommand{\tr}{\mathrm{Tr}}
\newcommand{\blockdiag}{\mathrm{blockdiag}}
\newcommand{\indep}{\stackrel{\mathrm{ind}}{\sim}}
\newcommand{\iid}{\stackrel{\mathrm{iid}}{\sim}}
\newcommand{\Bernoulli}{\mathrm{Bernoulli}}
\newcommand{\Betadist}{\mathrm{Beta}}
\newcommand{\BG}{\mathrm{BernoulliGraph}}
\newcommand{\PABM}{\mathrm{PABM}}
\newcommand{\RDPG}{\mathrm{RDPG}}
\newcommand{\GRDPG}{\mathrm{GRDPG}}
\newcommand{\Multinomial}{\mathrm{Multinomial}}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{proposition}{Proposition}
\theoremstyle{remark}
\newtheorem*{remark}{Remark}
\theoremstyle{definition}
\newtheorem{definition}{Definition}
\newtheorem{example}{Example}
\newpage
\spacingset{1.5} % DON'T change the spacing!

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

Statistical inference on random graphs requires a suitable probability model. 
A general probability model for unweighted and undirected graphs is the Bernoulli Graph (also known as the inhomogeneous Erd\"{o}s-R\'{e}nyi model), which assumes that edges occur as independent Bernoulli trials.  A Bernoulli Graph is characterized by an edge probability matrix $P=[P_{ij}]$, where an edge between vertices $i$ and $j$ occurs with success probability $P_{ij}$.  A trivial example of a Bernoulli Graph is the (homogeneous) Erd\"{o}s-R\'{e}nyi model proposed by \citet{Gilbert:1959}, in which the vertices of the random graph are fixed and possible edges occur independently with fixed probability $P_{ij}=p$.  The requirement that $p$ is fixed is too strong for most applications, and various researchers have weakened that requirement in various ways.  The present work relates two lines of generalization.

Network analysis is often concerned with community detection.  One form of community detection assumes that each vertex belongs to an unobserved community, with the probability of an edge between vertices $i$ and $j$ depending on the communities to which $i$ and $j$ belong.  Formally, one assigns each vertex $v_i$ a community label $z_i$ and assumes a Bernoulli Graph in which $P_{ij}$ is a function of $z_i$ and $z_j$.  Such models, called Block Models, define the goal of community detection as a problem in statistical inference: identify the true community (up to permutation of labels) to which each vertex belongs.

The classical Stochastic Block Model (SBM) of \citet{doi:10.1080/0022250X.1971.9989788} specifies that each edge probability $P_{ij}$ depends only on the labels $z_i$ and $z_j$, i.e., $P_{ij} = \omega_{z_,z_j}$.  Subsequent researchers have weakened this assumption.   The Degree-Corrected Block Model (DCBM) of \citet{Karrer_2011} assigns an additional parameter $\theta_i$ to each vertex and sets $P_{ij} = \theta_i \theta_j \omega_{z_,z_j}$. The Popularity Adjusted Block Model (PABM) of \citet{307cbeb9b1be48299388437423d94bf1} generalizes the DCBM, allowing heterogeneity of edge probabilities within and between communities while still maintaining distinct community structure.

Another type of Bernoulli Graph was proposed by
\citet*{10.1007/978-3-540-77004-6_11}.  A Random Dot Product Graph
(RDPG) specifies that each vertex corresponds to a latent position
vector in Euclidean space and that the probability of an edge between
two vertices is the dot product of their latent position vectors.
Thus, if the latent positions are $x_1,\ldots,x_n \in \mathbb{R}^d$
and $X = [\begin{array}{ccc} x_1 & \cdots & x_n \end{array} ]^\top$,
then the edge probability matrix is $P = XX^\top$.  Clearly, any
Bernoulli Graph with positive semidefinite $P$ is an RDPG.  The
positive definite Euclidean inner product in the RDPG model was
replaced by an indefinite inner product in
\citet{rubindelanchy2017statistical}, resulting in the {\em
Generalized}\/ RDPG (GRDPG).

In contrast to Block Models, neither RDPGs nor GRDPGs inherently
specify distinct communities.  However, one can easily impose
community structure by assuming that the latent positions lie in
distinct clusters.  Hence, it is not surprising that Block Models can
be studied by reformulating them as RDPGs or GRDPGs.  For example, an
assortative SBM (an SBM for which $P$ is positive semidefinite) is
equivalent to an RDPG for which all vertices in the same community
correspond to the same latent position vector.
Likewise, the DCBM is equivalent to an RDPG for which all vertices in
the same community correspond to latent position vectors that lie on a
straight line. %(\cite{lyzinski2014, rubindelanchy2017statistical}).

Because the edge probability matrix of a PABM is not necessarily
positive semidefinite, a PABM is not necessarily an RDPG.  In Section
\ref{connecting-the-pabm-to-the-grdpg} we demonstrate that every PABM
is in fact a specific type of GRDPG for which the latent position
vectors lie in distinct orthogonal subspaces, each subspace
corresponding to a community.  This identification is our central
result.  In Section \ref{methods}, we use the geometry of the GRDPG to
derive more efficient algorithms for detecting the communities and
estimating the parameters in the PABM.  We report the results of
simulation studies in Section \ref{simulated-examples} and apply our
methods to three well-known data sets in Section
\ref{real-data-examples}. Section \ref{discussion} concludes.

\hypertarget{connecting-the-popularity-adjusted-block-model-to-the-generalized-random-dot-product-graph}{%
\section{PABMs are GRDPGs}\label{connecting-the-popularity-adjusted-block-model-to-the-generalized-random-dot-product-graph}}

In this section, we show that the PABM is a special case of the GRDPG. 
More specifically, a graph $G$ drawn from the PABM can be represented by 
a collection of latent vectors in Euclidean space. 
We further show that the latent configuration that
induces the PABM consists of orthogonal subspaces with each subspace
corresponding to a community.

\hypertarget{notation}{%
\subsection{Notation and Scope}\label{notation}}

Let $G = (V, E)$ be an unweighted, undirected, 
and hollow graph with vertex set $V$ ($|V| = n$) and edge set $E$. The matrix
$A \in \{0, 1\}^{n \times n}$ represents the adjacency matrix of $G$ 
such that $A_{ij} = 1$ if there exists an edge between vertices $i$ and $j$ 
and $0$ otherwise. Because $G$ is symmetric and hollow,
$A_{ij} = A_{ji}$ and $A_{ii} = 0$ for each $i, j \in [n]$. 
We further restrict our analyses to Bernoulli graphs. 
Let $P \in [0, 1]^{n \times n}$ be a symmetric matrix of edge probabilities. 
Graph $G$ is sampled from $P$ by drawing $A_{ij} \indep \Bernoulli(P_{ij})$ 
for each $1 \leq i < j \leq n$ (setting $A_{ji} = A_{ij}$ and $A_{ii} = 0$). 
We denote $A \sim \BG(P)$ as a graph with adjacency matrix $A$ 
sampled from edge probability matrix $P$ in this manner. 
If each vertex has a hidden label in $[K]$, 
they are denoted as $z_1, ..., z_n$. 
Finally, we denote 
$X = \bigl[x_1 \mid \cdots \mid x_n\bigr]^\top \in \mathbb{R}^{n \times d}$ 
as the matrix corresponding to a collection of $n$ latent vectors $x_1, ..., x_n \in \mathbb{R}^d$.

\hypertarget{the-popularity-adjusted-block-model-and-the-generalized-random-dot-product-graph}{%
\subsection{Two Probability Models for Graphs}\label{the-popularity-adjusted-block-model-and-the-generalized-random-dot-product-graph}}

\begin{definition}[Popularity Adjusted Block Model]
\label{pabm}
Let $P \in [0, 1]^{n \times n}$ be a symmetric edge probability matrix for a
graph $G = (V, E)$ with adjacency matrix $A$ 
such that $A \sim \mathrm{Bernoulli}(P)$.
Let each vertex have a community label between $1$ and $K$.
Then $G$ is drawn from a Popularity Adjusted Block Model if
each vertex has $K$ popularity parameters that describe its affinity
toward each of the $K$ communities, i.e., vertex $i$ has popularity parameters
$\lambda_{i1}, ..., \lambda_{iK}$, and  $P_{ij} = \lambda_{i z_j}
\lambda_{j z_i}$ for all $i \leq j$.

Another characterization of the PABM is as follows.
Let the rows and columns of $P$ be arranged by community label
such that $n_k \times n_{\ell}$ block $P^{(k \ell)}$
describes the edge probabilities between vertices in communities
$k$ and $\ell$; note that $P^{(k \ell)} = (P^{(k \ell)})^\top$.
Now suppose that each block $P^{(k \ell)}$ can be written as the outer product of two vectors:
\begin{equation} \label{eq:pabm}
  P^{(k \ell)} = \lambda^{(k \ell)} (\lambda^{(\ell k)})^{\top}
\end{equation}
for a set of $K^2$ popularity vectors 
$\{\lambda^{(st)}\}_{1 \leq s \leq K, 1 \leq t \leq K}$ where each
$\lambda^{(st)}$ is a column vector of dimension $n_s$. Then a graph $G$
is drawn from a PABM with parameters
$\{\lambda^{(st)}\}$ if its adjacency matrix $A$
satisfies $A \sim \mathrm{BernoulliGraph}(P)$.
\end{definition}

We will use the notation \(A \sim \PABM(\{\lambda^{(k \ell)}\}_K)\) to denote
a random adjacency matrix \(A\) drawn from a PABM with parameters
\(\{\lambda^{(k \ell)}\}\) consisting of \(K\) underlying communities.

\begin{definition}[Generalized Random Dot Product Graph]
\label{grdpg}
Let graph $G = (V, E)$ be drawn as $A \sim \BG(P)$ for some symmetric $P \in
[0,1]^{n \times n}$. 
If there exists a data matrix $X \in \mathbb{R}^{n \times d}$ such that
\begin{equation} \label{eq:grdpg}
  P = X I_{p,q} X^\top
\end{equation}
for some $d, p, q \in \mathbb{N}$ and $p + q = d$, then
$G$ is drawn from the Generalized Random Dot Product Graph 
with latent positions $x_1, ..., x_n \in \mathbb{R}^d$ and signature $(p, q)$.
\end{definition}

We will use the notation \(A \sim \GRDPG_{p,q}(X)\) to denote a random
adjacency matrix \(A\) drawn from latent positions \(X\) and signature
\((p, q)\). If instead of fixed latent positions, 
they are drawn $X_1, ..., X_n \iid F$, 
we denote the GRDPG as $(A, X) \sim \GRDPG_{p,q}(F, n)$.

\begin{remark}
We can use Adjacency Spectral Embedding (ASE) 
\citep{doi:10.1080/01621459.2012.699795} to recover the latent vectors of a GRDPG. 
This procedure consists of taking the spectral decomposition of $A$ 
(or $P$ if available) and choosing the $p$ most positive and $q$ most negative
eigenvalues and their corresponding eigenvectors 
to construct the embedding $\hat{Z} = V |D|^{1/2}$. 
\end{remark}

\begin{definition}[Indefinite Orthogonal Group]
The indefinite orthogonal group with signature $(p, q)$ is
the set $\{Q \in \mathbb{R}^{d \times d} \colon Q I_{p, q} Q^{\top} = I_{p, q}\}$,
denoted as $\mathbb{O}(p, q)$. Here $d = p + q$. 
\end{definition}

\begin{remark}
The latent vectors that produce $X I_{p,q} X^\top = P$ are not unique
\citep{rubindelanchy2017statistical}.
More specifically, if $P_{ij} = x_i^\top I_{p, q} x_j$, 
then we also have for any $Q \in \mathbb{O}(p, q)$ that
$(Q x_i)^\top I_{p, q} (Q x_j) = x_i^\top (Q^\top I_{p, q} Q) x_j =
x_i^\top I_{p, q} x_j = P_{ij}$.
Unlike in the RDPG case, transforming the latent positions via multiplication
by $Q \in \mathbb{O}(p, q)$ does not necessarily maintain interpoint angles or
distances.
\end{remark}

\hypertarget{connecting-the-pabm-to-the-grdpg}{%
\subsection{The Geometry of PABMs}\label{connecting-the-pabm-to-the-grdpg}}

Now that we defined the PABM and GRDPG, 
we show the special geometry of the PABM when viewed as a GRDPG.

\begin{theorem}[The latent configuration of the PABM]
\label{theorem2}
Let $A \sim \mathrm{PABM}(\{\lambda^{(k \ell)}\}_K)$ be an instance of a
PABM with $K \geq 1$ blocks and latent vectors $\{\lambda^{(k \ell)}
\colon 1 \leq k \leq K, 1 \leq \ell \leq K\}$. 
Then there exists a block diagonal matrix
$X \in \mathbb{R}^{n \times K^2}$ defined by $\{\lambda^{(k \ell)}\}$ and a 
$K^2 \times K^2$ {\em fixed} orthonormal matrix $U$ such 
that $A \sim \mathrm{GRDPG}_{K (K+1) / 2, K (K-1) / 2}(XU)$.
\end{theorem}

\begin{proof}
We will prove this theorem in two parts. First, for demonstration purposes, we focus on the case for $K = 2$. Then we generalize this to $K \geq 2$.

For the $K = 2$ case, the proof is straightforward. Let 
$$X = \begin{bmatrix}
\lambda^{(11)} & \lambda^{(12)} & 0 & 0 \\
0 & 0 & \lambda^{(21)} & \lambda^{(22)}
\end{bmatrix} \quad \text{and} \quad
U = \begin{bmatrix} 1 & 0 & 0 & 0 \\
0 & 0 & 1 / \sqrt{2} & 1 / \sqrt{2} \\
0 & 0 & 1 / \sqrt{2} & - 1 / \sqrt{2} \\
0 & 1 & 0 & 0 \end{bmatrix}.$$
Then by straightforward matrix multiplication, we obtain 
\[X U I_{3, 1} U^\top X^\top =
\begin{bmatrix}
  \lambda^{(11)} (\lambda^{(11)})^\top & \lambda^{(12)} (\lambda^{(21)})^\top \\
  \lambda^{(21)} (\lambda^{(12)})^\top & \lambda^{(22)} (\lambda^{(22)})^\top
\end{bmatrix}, \]
which is a GRDPG with latent vectors described by $X U$.

It is nevertheless instructive to look at a few intermediate steps. 
More specifically, the product $U I_{3, 1} U^\top$ 
yields a permutation matrix $\Pi$ with fixed points at positions $1$ and $4$ 
and a cycle of order 2 swapping positions $2$ and $3$, i.e., 
$$\Pi = U I_{3, 1} U^\top = \begin{bmatrix} 1 & 0 & 0 & 0 \\
  0 & 0 & 1 & 0 \\
  0 & 1 & 0 & 0 \\
  0 & 0 & 0 & 1
\end{bmatrix}.$$
Furthermore, as $U$ is orthonormal and $I_{3, 1}$ is diagonal, 
$U I_{3, 1} U^\top$ is also an eigendecomposition of $\Pi$ where the fixed
points of $\Pi$ are mapped to the eigenvectors $e_1$ and $e_4$
while the cycles of order two are mapped to the eigenvectors  
$\tfrac{1}{\sqrt{2}}(e_{2} + e_3)$ and $\tfrac{1}{\sqrt{2}}(e_{2} -
e_3)$; here $e_i$ denote the $i^\mathrm{th}$ basis vector in $\mathbb{R}^{4}$.

For the general case, we first define the following matrices
\begin{gather}
\label{eq:xy}
\Lambda^{(k)} = \begin{bmatrix} \lambda^{(k1)} \mid \cdots \mid \lambda^{(kK)} \end{bmatrix}
\in \mathbb{R}^{n_k \times K}, \quad
X = \text{blockdiag}(\Lambda^{(1)}, \dots, \Lambda^{(K)}) \in
\mathbb{R}^{n \times K^2}, \\
L^{(k)} = \text{blockdiag}(\lambda^{(1k)}, \dots, \lambda^{(Kk)}) \in
\mathbb{R}^{n \times K}, \quad
Y = \begin{bmatrix} L^{(1)} \mid \cdots \mid L^{(K)} \end{bmatrix} \in
\mathbb{R}^{n \times K^2}.
\end{gather}
It is then straighforward to verify that
\begin{gather*}
  XY^{\top} = \text{blockdiag}(\Lambda^{(1)}, \dots,
\Lambda^{(K)}) \begin{bmatrix} L_1^{\top} \\ \vdots \\
  L_{K}^{\top} \end{bmatrix} = \begin{bmatrix} \Lambda^{(1)}
  (L^{(1)})^{\top} \\ \vdots \\
  \Lambda^{(K)} (L^{(K)})^{\top} \end{bmatrix}, \\
\Lambda^{(k)} (L^{(k)})^{\top} = \begin{bmatrix} \lambda^{(k1)}
  (\lambda^{(1k)})^{\top} \mid \dots \mid \lambda^{(kK)}
  (\lambda^{(Kk)})^{\top} \end{bmatrix} = \begin{bmatrix} P^{(k1)}
  \mid P^{(k2)} \mid \dots \mid P^{(kK)} \end{bmatrix}.
\end{gather*}
We therefore have $P = X Y^\top$. 
Similar to the $K = 2$ case, we also have $Y = X \Pi$ for some permutation matrix
$\Pi$ and hence $P = X \Pi X^\top$.
The permutation described by $\Pi$ has $K$ fixed points, which correspond to
$K$ eigenvalues equal to $1$ with corresponding eigenvectors $e_k$ where
$k = r (K + 1) + 1$ for $r = 0, ..., K - 1$. It also has
$\binom{K}{2} = K (K - 1) / 2$ cycles of order $2$. Each cycle corresponds to
a pair of eigenvalues $\{-1,+1\}$ and a pair of eigenvectors
$\{(e_s + e_t)/\sqrt{2},(e_s - e_t)/ \sqrt{2}\}$.

Let $p = K (K + 1) / 2$ and $q = K (K - 1) / 2$. 
We therefore have
\begin{equation} \label{eq:permutation}
\Pi = U I_{p,q} U^\top
\end{equation}
where $U$ is a $K^2 \times K^2$ orthogonal matrix and hence
\begin{equation} \label{eq:pabm-grdpg}
P = X U I_{p, q} (X U)^\top.
\end{equation}
In summary we can describe the PABM with $K$ communities as a GRDPG with latent
positions $X U$ and signature $(p,q) = \bigl( \tfrac{1}{2} K (K + 1) ,
\tfrac{1}{2} K (K - 1)\bigr)$.
\end{proof}

\begin{example} Let $A$ be a $3$ blocks PABM with latent vectors
  $\{\lambda^{(k \ell)} \colon 1 \leq k \leq 3, 1 \leq \ell \leq 3\}$. Using the same notation as in Theorem
  \ref{theorem2}, we can define
\begin{gather*}
X = \begin{bmatrix}
\lambda^{(11)} & \lambda^{(12)} & \lambda^{(13)} & 0 & 0 & 0 & 0 & 0 & 0 \\
0 & 0 & 0 & \lambda^{(21)} & \lambda^{(22)} & \lambda^{(23)} & 0 & 0 & 0 \\
0 & 0 & 0 & 0 & 0 & 0 & \lambda^{(31)} & \lambda^{(32)} & \lambda^{(33)}
\end{bmatrix}, \\
Y = \begin{bmatrix}
\lambda^{(11)} & 0 & 0 & \lambda^{(12)} & 0 & 0 & \lambda^{(13)} & 0 & 0 \\
0 & \lambda^{(21)} & 0 & 0 & \lambda^{(22)} & 0 & 0 & \lambda^{(23)} & 0 \\
0 & 0 & \lambda^{(31)} & 0 & 0 & \lambda^{(32)} & 0 & 0 & \lambda^{(33)}
\end{bmatrix}.
\end{gather*}
Then $Y = X \Pi$ and $P = X Y^{\top}$ where $\Pi$ is a $9 \times 9$ 
permutation matrix of the form
$$\Pi = 
\Bigl[e_1 \mid e_4 \mid e_7 \mid e_2 \mid e_5 \mid e_8 \mid e_3
\mid e_6 \mid e_9 \Bigr].$$
where $e_i$ denotes the $i^{th}$ basis vector in $\mathbb{R}^{9}$. 
The matrix $\Pi$ corresponds to a permutation of $\{1,2,\dots,9\}$
with the following decomposition.
\begin{enumerate}
\item Positions 1, 5, 9 are fixed.
\item There are three cycles of length 2, namely $(2, 4)$, $(3, 7)$, and $(6, 8)$.
\end{enumerate}
We can therefore write $\Pi$ as $\Pi = U I_{6, 3} U^\top$ where the first three
columns of $U$ consist of $e_1$, $e_5$, and $e_9$ corresponding to the
{\em fixed} points, the next three columns consist of eigenvectors
$(e_k + e_{\ell}) / \sqrt{2}$, and the last three columns consist of eigenvectors
$(e_k - e_{\ell}) / \sqrt{2}$ for $(k, \ell) \in
\{(2,4),(3,7),(6,8)\}$.

The matrix $P$ is then the edge probabilities matrix for a 
Generalized Random Dot Product Graph whose latent positions 
are the rows of the matrix
$$XU = \begin{bmatrix}
  \lambda^{(11)} & 0 & 0 &
  \frac{\lambda^{(12)}}{\sqrt{2}} & \frac{\lambda^{(13)}}{\sqrt{2}} & 0 &
  \frac{\lambda^{(12)}}{\sqrt{2}} & \frac{\lambda^{(13)}}{\sqrt{2}} & 0 \\
  0 & \lambda^{(22)} & 0 &
  \frac{\lambda^{(21)}}{\sqrt{2}} & 0 & \frac{\lambda^{(23)}}{\sqrt{2}} &
  -\frac{\lambda^{(21)}}{\sqrt{2}} & 0 & \frac{\lambda^{(23)}}{\sqrt{2}} \\
  0 & 0 & \lambda^{(33)} &
  0 & \frac{\lambda^{(31)}}{\sqrt{2}} & \frac{\lambda^{(32)}}{\sqrt{2}} &
  0 & -\frac{\lambda^{(31)}}{\sqrt{2}} & -\frac{\lambda^{(32)}}{\sqrt{2}}
\end{bmatrix}.$$
\end{example}

\hypertarget{methods}{%
\section{Algorithms}\label{methods}}

Two inference objectives arise from the PABM:

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames*,x11names*}{xcolor}
%
\documentclass[
  12pt,
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Supplemental Materials for ``Popularity Adjusted Block Models are Generalized Random Dot Product Graphs''},
  colorlinks=true,
  linkcolor=Maroon,
  filecolor=Maroon,
  citecolor=Blue,
  urlcolor=blue,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\usepackage{float}
\usepackage{mathtools}
\usepackage{natbib}
\usepackage[linesnumbered,ruled,vlined]{algorithm2e}
\setcitestyle{numbers,square,comma}
\usepackage{verbatim}
\usepackage{amsthm}
\usepackage{comment}
\ifluatex
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{plainnat}

\title{Supplemental Materials for ``Popularity Adjusted Block Models are
Generalized Random Dot Product Graphs''}
\author{}
\date{\vspace{-2.5em}}

\begin{document}
\maketitle

\newcommand{\diag}{\mathrm{diag}}
\newcommand{\tr}{\mathrm{Tr}}
\newcommand{\blockdiag}{\mathrm{blockdiag}}
\newcommand{\indep}{\stackrel{\mathrm{ind}}{\sim}}
\newcommand{\iid}{\stackrel{\mathrm{iid}}{\sim}}
\newcommand{\Bernoulli}{\mathrm{Bernoulli}}
\newcommand{\Betadist}{\mathrm{Beta}}
\newcommand{\BG}{\mathrm{BernoulliGraph}}
\newcommand{\PABM}{\mathrm{PABM}}
\newcommand{\RDPG}{\mathrm{RDPG}}
\newcommand{\GRDPG}{\mathrm{GRDPG}}
\newcommand{\Multinomial}{\mathrm{Multinomial}}

\hypertarget{sparsity-simulation-study}{%
\section{Sparsity Simulation Study}\label{sparsity-simulation-study}}

In this simulation study, we use the same setup as in the simulations
for balanced communities but with fixed \(K = 3\) and \(n = 2^{11}\) and
vary the sparsity parameter \(\rho \in (0, 1]\) (this was fixed at
\(\rho_n = 1\) for the previous simulations. More specifically,

\begin{itemize}
\tightlist
\item
  Number of vertices \(n = 2048\)
\item
  Number of underlying communities \(K = 3\)
\item
  Sparsity parameter \(\rho_n = 0.1, 0.3, 0.5, 0.7, 0.9\)
\item
  Mixture parameters \(\alpha_k = 1 / K\) for \(k = 1, ..., K\) (i.e.,
  each community label has an equal probability of being drawn)
\item
  Community labels
  \(z_k \stackrel{\mathrm{iid}}{\sim}\mathrm{Multinomial}(\alpha_1, ..., \alpha_K)\)
\item
  Within-group popularities
  \(\lambda^{(kk)} \stackrel{\mathrm{iid}}{\sim}\mathrm{Beta}(2, 1)\)
\item
  Between-group popularities
  \(\lambda^{(k \ell)} \stackrel{\mathrm{iid}}{\sim}\mathrm{Beta}(1, 2)\)
  for \(k \neq \ell\)
\end{itemize}

Fifty simulations were performed for each combination of \(n\) and
\(K\). The results for community detection and parameter estimation
error are shown in Fig. \ref{fig:sparsity_sim}. Community detection
error is defined as the number of misclustered vertices. Parameter
estimation error is defined as \(\frac{1}{n} \|P - \hat{P}\|_F\) where
\(\hat{P}\) is the estimated edge probability matrix.
